---
title: "Visualizing Plastics: Deliverable from Data"
author: "Nathan Lee, Noah Nayve, Ryan Ekdahl, Taiga Sobajima"
format:
  dashboard:
    theme: flatly
    logo: "logo.png"
    nav-buttons:
      - icon: envelope
        href: mailto:nlee174@calpoly.edu
        aria-label: Nathan Lee
      - icon: envelope
        href: mailto:nnayve@calpoly.edu
        aria-label: Noah Nayve
      - icon: envelope
        href: mailto:rekdahl@calpoly.edu
        aria-label: Ryan Ekdahl
      - icon: envelope
        href: mailto:tsobajim@calpoly.edu
        aria-label: Taiga Sobajima
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(glue)
library(gt)
library(plotly)
library(scales)
library(ggridges)
library(broom)
library(ggrepel)
```

```{r}
#| label: load-data
#| message: false
plastics <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
population_plastics <- read_csv("data/population_plastics.csv")
```

```{r}
#| label: functions
#| message: false
plastic_volunteer_summary <- function(polyethylene, country_name, data = plastics) {
  if (!(polyethylene %in% c("ldpe", "hdpe"))) {
    stop(paste("Plastic type not found:", polyethylene))
  }
  if (!(country_name %in% data$country)) {
    stop(paste("Country not found:", country_name))
  }
  caps_dpe <- toupper(polyethylene)
  avg_col <- glue("Ave_{caps_dpe}")
  percent_col <- glue("Percent_{caps_dpe}")
  total_col <- glue("Total_Plastic")
  result <- data |>
    filter(country == country_name) |>
    mutate(percent_plastic_type = .data[[polyethylene]] / grand_total) |>
    summarise(
      !!avg_col := mean(.data[[polyethylene]], na.rm = TRUE),
      !!percent_col := sum(percent_plastic_type, na.rm = TRUE) /
                        sum(!is.na(percent_plastic_type)) * 100,
      !!total_col := sum(grand_total, na.rm = TRUE),
      Ave_Vol = first(volunteers) / n(),
      Total_Vol = first(volunteers)
    ) |>
    mutate(across(where(is.numeric), ~round(., 2)),
           !!percent_col := paste0(.data[[percent_col]], "%"))
  return(result)
}

summarize_plastic_and_ecology <- function(country_name, data = population_plastics) {
  if (!(country_name %in% data$country)) {
    stop(paste("Country not found:", country_name))
  }
  country_stats <- data |>
    filter(country == country_name, parent_company == "Grand Total", grand_total > 0) |>
    summarize(
      country = country_name,
      avg_pet = mean(pet / grand_total, na.rm = TRUE),
      avg_hdpe = mean(hdpe / grand_total, na.rm = TRUE),
      avg_ldpe = mean(ldpe / grand_total, na.rm = TRUE),
      avg_pp = mean(pp / grand_total, na.rm = TRUE),
      avg_ps = mean(ps / grand_total, na.rm = TRUE),
      avg_pvc = mean(pvc / grand_total, na.rm = TRUE),
      avg_o = mean(o / grand_total, na.rm = TRUE),
      avg_forested = mean(forested_area, na.rm = TRUE),
      avg_threatened = mean(threatened_species, na.rm = TRUE)
    )
  leading_plastics <- country_stats |>
    pivot_longer(cols = !c(country, avg_forested, avg_threatened),
                 names_to = "plastic_type", values_to = "avg_value") |>
    slice_max(order_by = avg_value, n = 1, with_ties = FALSE) |>
    mutate(plastic_type = str_remove(plastic_type, "avg_"),
           plastic_type = str_to_upper(plastic_type),
           plastic_type = if_else(plastic_type == "O", "Other", plastic_type)) |>
    pull(plastic_type)
  country_stats |> mutate(leading_plastic = leading_plastics)
}

all_countries_rq2 <- population_plastics |>
  filter(parent_company == "Grand Total", grand_total > 0,
         !is.na(forested_area), !is.na(threatened_species)) |>
  distinct(country) |>
  pull(country)

plastic_ecology_summary_table <- map(all_countries_rq2, summarize_plastic_and_ecology) |>
  bind_rows()
```

# About

## Row

### 

**Visualizing Plastics: Deliverable from Data**

This dashboard presents findings from an analysis of global plastic waste collection data from the Break Free from Plastic (BFFP) audit, covering 2019 and 2020 across 45+ countries.

**Research Questions:**

1. Do countries with more volunteers tend to collect more proportions of both high and low density polyethylene plastic in 2019 and 2020?

2. Do countries with greater forest coverage and more threatened species show a different breakdown of plastic types?

**Data Sources:**

- Main dataset: Break Free from Plastic audit data (TidyTuesday)
- Supplementary dataset: API Ninjas Country API

### 

**Key Findings**

- No clear relationship between volunteer count and polyethylene proportion was found, though a general positive trend exists in Figure 1.

- PET plastic drops dramatically as biodiversity increases — from 71.7% in low biodiversity countries to 21.5% in high biodiversity countries.

- Threatened species count is the only statistically significant predictor of PET proportion (p = 0.021), explaining 12.5% of variance.

# Summaries & Visualizations

## Row

### Research Question 1 {.tabset}

#### Tables

```{r}
top5_countries <- plastics |>
  group_by(country) |>
  summarise(ave_vol = first(volunteers) / n()) |>
  slice_max(ave_vol, n = 5) |>
  pull(country)

bot5_countries <- plastics |>
  group_by(country) |>
  summarise(ave_vol = first(volunteers) / n()) |>
  slice_min(ave_vol, n = 5) |>
  pull(country)

top5_hdpe <- map(top5_countries, ~plastic_volunteer_summary("hdpe", .x) |>
                   mutate(Country = .x, .before = 1)) |> list_rbind()
bot5_hdpe <- map(bot5_countries, ~plastic_volunteer_summary("hdpe", .x) |>
                   mutate(Country = .x, .before = 1)) |> list_rbind()

bind_rows(
  top5_hdpe |> mutate(Group = "Top 5", .before = 2),
  bot5_hdpe |> mutate(Group = "Bottom 5", .before = 2)
) |>
  mutate(across(where(is.numeric), ~ifelse(is.nan(.), 0, .)),
         across(where(is.character), ~ifelse(. == "NaN%", "0%", .))) |>
  gt() |>
  tab_header(title = md("**HDPE Summary**"),
             subtitle = "Top and Bottom 5 Average Volunteer Countries") |>
  tab_caption("Table 1: HDPE statistics for top and bottom 5 volunteer countries") |>
  tab_spanner(label = "Location", columns = c(Country, Group)) |>
  tab_spanner(label = "HDPE Metrics", columns = c(Ave_HDPE, Percent_HDPE, Total_Plastic)) |>
  tab_spanner(label = "Volunteer Metrics", columns = c(Ave_Vol, Total_Vol)) |>
  cols_label(Ave_HDPE = "Avg HDPE", Percent_HDPE = "% HDPE",
             Total_Plastic = "Total Plastic", Ave_Vol = "Avg Volunteers",
             Total_Vol = "Total Volunteers") |>
  tab_style(style = cell_fill(color = "#457B9D"), locations = cells_column_labels()) |>
  tab_style(style = cell_text(color = "white", weight = "bold"), locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#1D3557"), locations = cells_column_spanners()) |>
  tab_style(style = cell_text(color = "white", weight = "bold"), locations = cells_column_spanners())
```

#### Visualizations

```{r}
#| fig-alt: "Bubble plot showing average volunteers vs HDPE percentage by country"
all_countries_rq1 <- plastics |> distinct(country) |> pull(country)

bubble_data <- map(all_countries_rq1, ~plastic_volunteer_summary("hdpe", .x) |>
                     mutate(Country = .x)) |>
  list_rbind() |>
  mutate(Percent_HDPE = as.numeric(str_remove(Percent_HDPE, "%"))) |>
  filter(Percent_HDPE < 100)

bubble_plot <- ggplot(bubble_data, aes(x = Ave_Vol, y = Percent_HDPE,
                        size = Total_Plastic, color = Country,
                        text = paste("Country:", Country,
                                     "<br>Avg Volunteers:", Ave_Vol,
                                     "<br>% HDPE:", Percent_HDPE,
                                     "<br>Total Plastic:", comma(Total_Plastic)))) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(range = c(2, 15)) +
  scale_x_log10() +
  scale_color_viridis_d() +
  guides(color = "none") +
  theme_minimal() +
  labs(title = "Figure 1: Average Volunteers vs. HDPE Plastic (%) by Country",
       subtitle = "Bubble size represents total plastic collected",
       x = "Average Volunteers (log scale)", y = "", size = "Total Plastic")

ggplotly(bubble_plot, tooltip = "text")
```

### Research Question 2 {.tabset}

#### Tables

```{r}
plastic_ecology_summary_table |>
  mutate(forest_coverage_level = cut(avg_forested,
                            breaks = c(0, 19, 36, 100),
                            labels = c("Low (< 19%)", "Medium (19-36%)", "High (> 36%)"))) |>
  group_by(forest_coverage_level) |>
  slice_min(avg_forested, n = 5) |>
  ungroup() |>
  arrange(avg_forested) |>
  gt(groupname_col = "forest_coverage_level") |>
  tab_caption("Table 2: Plastic type proportions by forest coverage level") |>
  tab_header(title = "Plastic Type Proportions by Country and Forest Coverage",
             subtitle = "Top 5 least forested countries within each coverage level") |>
  cols_label(country = "Country", avg_pet = "PET", avg_hdpe = "HDPE",
             avg_ldpe = "LDPE", avg_pp = "PP", avg_ps = "PS", avg_pvc = "PVC",
             avg_o = "Other", avg_forested = "Forested Area (%)",
             avg_threatened = "Threatened Species", leading_plastic = "Leading Plastic") |>
  cols_move(columns = leading_plastic, after = country) |>
  tab_spanner(label = "Average Plastic Proportion", columns = avg_pet:avg_o) |>
  fmt_percent(columns = avg_pet:avg_o, decimals = 1) |>
  fmt_number(columns = avg_forested, decimals = 1) |>
  fmt_number(columns = avg_threatened, decimals = 0) |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_spanners()) |>
  tab_style(style = cell_fill(color = "#18BC9C"), locations = cells_row_groups(groups = "High (> 36%)")) |>
  tab_style(style = cell_fill(color = "#F39C12"), locations = cells_row_groups(groups = "Medium (19-36%)")) |>
  tab_style(style = cell_fill(color = "#E74C3C"), locations = cells_row_groups(groups = "Low (< 19%)")) |>
  tab_style(style = cell_text(weight = "bold", color = "#2C3E50"),
            locations = cells_body(columns = leading_plastic)) |>
  opt_stylize(style = 6, color = "gray") |>
  tab_options(column_labels.background.color = "#2C3E50",
              heading.background.color = "#2C3E50") |>
  sub_missing(missing_text = "—") |>
  cols_hide(forest_coverage_level)
```

#### Visualizations

```{r}
#| message: false
#| warning: false
#| fig-alt: "Scatter plot showing forested area vs threatened species colored by leading plastic type"
plot_data <- plastic_ecology_summary_table |>
  mutate(leading_plastic = fct_infreq(leading_plastic) |> fct_relevel("Other", after = Inf),
         point_alpha = if_else(leading_plastic == "Other", 0.4, 0.9))

ggplot(plot_data, aes(x = avg_forested, y = avg_threatened,
                      color = leading_plastic, label = country)) +
  geom_point(aes(alpha = point_alpha), size = 1.5) +
  geom_text_repel(size = 2, max.overlaps = Inf, box.padding = 0.25,
                  point.padding = 0.15, segment.color = "gray70",
                  segment.size = 0.25, show.legend = FALSE) +
  scale_y_log10(labels = scales::label_comma()) +
  scale_alpha_identity() +
  scale_color_manual(values = c("PET" = "#E74C3C", "HDPE" = "#3498DB",
                                "LDPE" = "#F39C12", "PP" = "#18BC9C",
                                "PS" = "#9B59B6", "PVC" = "#2C3E50",
                                "Other" = "#95A5A6")) +
  labs(x = "Forested Area (%)", y = "Threatened Species Count (log scale)",
       color = "Leading Plastic",
       title = "Figure 3: Where Do Different Plastic Types Dominate?") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))
```

# Data Model

## Row

### Model Table

```{r}
pet_model <- population_plastics |>
  filter(parent_company == "Grand Total", grand_total > 0,
         !is.na(forested_area), !is.na(threatened_species),
         !is.na(gdp_per_capita), !is.na(volunteers)) |>
  mutate(pet_prop = pet / grand_total) |>
  lm(pet_prop ~ forested_area + log(threatened_species) +
                log(gdp_per_capita) + volunteers, data = _)

anova(pet_model) |>
  tidy() |>
  mutate(pct_variance = sumsq / sum(sumsq) * 100,
         term = recode(term, "volunteers" = "Volunteers",
                       "forested_area" = "Forested Area",
                       "log(threatened_species)" = "Threatened Species (log)",
                       "log(gdp_per_capita)" = "GDP Per Capita (log)",
                       "Residuals" = "Unexplained")) |>
  select(term, df, sumsq, pct_variance, statistic, p.value) |>
  gt() |>
  tab_caption("Table 3: ANOVA results predicting PET plastic proportion") |>
  tab_header(title = "What Predicts PET Plastic Proportion?",
             subtitle = "Variance in PET proportion partitioned across predictors") |>
  cols_label(term = "Source", df = "df", sumsq = "Sum of Squares",
             pct_variance = "% of Variance", statistic = "F", p.value = "p-value") |>
  fmt_number(columns = c(sumsq, statistic), decimals = 3) |>
  fmt_number(columns = pct_variance, decimals = 1) |>
  fmt_number(columns = p.value, decimals = 4) |>
  sub_missing(missing_text = "—") |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) |>
  tab_style(style = list(cell_fill(color = "#18BC9C"), cell_text(weight = "bold")),
            locations = cells_body(rows = p.value < 0.05)) |>
  tab_options(column_labels.background.color = "#2C3E50",
              heading.background.color = "#2C3E50") |>
  tab_source_note("Note: PET modeled as it showed strongest relationship with ecological variables.") |>
  tab_source_note(paste0("Model R² = ", round(summary(pet_model)$r.squared, 3),
                         " · n = ", nobs(pet_model)))
```

### Model Interpretation 

**Key Finding**

Threatened species count is the only statistically significant predictor of PET plastic proportion (F = 5.841, p = 0.021), explaining 12.5% of variance. As threatened species increases, PET proportion decreases (coefficient = -0.1325).

Forested area showed a similar negative trend (p = 0.079) but was not statistically significant. Together the model explains about 23% of variation in PET proportion, suggesting that ecological richness is meaningfully associated with plastic composition patterns across countries.