---
title: "Visualizing Plastics: Deliverable from Data"
author: "Nathan Lee, Noah Nayve, Ryan Ekdahl, Taiga Sobajima"
format:
  dashboard:
    theme: flatly
    logo: "logo.png"
    nav-buttons:
      - icon: envelope
        href: mailto:nlee174@calpoly.edu
        aria-label: Nathan Lee
      - icon: envelope
        href: mailto:nnayve@calpoly.edu
        aria-label: Noah Nayve
      - icon: envelope
        href: mailto:rekdahl@calpoly.edu
        aria-label: Ryan Ekdahl
      - icon: envelope
        href: mailto:tsobajim@calpoly.edu
        aria-label: Taiga Sobajima
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(glue)
library(gt)
library(plotly)
library(scales)
library(ggridges)
library(broom)
library(ggrepel)
```

```{r}
#| label: load-data
#| message: false
plastics <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
population_plastics <- read_csv("data/population_plastics.csv")
```

```{r}
#| label: prep
#| message: false
#| warning: false
plastic_volunteer_summary <- function(polyethylene, country_name, data = plastics) {
  if (!(polyethylene %in% c("ldpe", "hdpe"))) stop(paste("Plastic type not found:", polyethylene))
  if (!(country_name %in% data$country)) stop(paste("Country not found:", country_name))
  caps_dpe <- toupper(polyethylene)
  avg_col <- glue("Ave_{caps_dpe}")
  percent_col <- glue("Percent_{caps_dpe}")
  result <- data |>
    filter(country == country_name) |>
    mutate(percent_plastic_type = .data[[polyethylene]] / grand_total) |>
    summarise(
      !!avg_col := mean(.data[[polyethylene]], na.rm = TRUE),
      !!percent_col := sum(percent_plastic_type, na.rm = TRUE) /
                        sum(!is.na(percent_plastic_type)) * 100,
      Total_Plastic = sum(grand_total, na.rm = TRUE),
      Ave_Vol = first(volunteers) / n(),
      Total_Vol = first(volunteers)
    ) |>
    mutate(across(where(is.numeric), ~round(., 2)),
           !!percent_col := paste0(.data[[percent_col]], "%"))
  return(result)
}

summarize_plastic_and_ecology <- function(country_name, data = population_plastics) {
  if (!(country_name %in% data$country)) stop(paste("Country not found:", country_name))
  country_stats <- data |>
    filter(country == country_name, parent_company == "Grand Total", grand_total > 0) |>
    summarize(
      country = country_name,
      avg_pet = mean(pet / grand_total, na.rm = TRUE),
      avg_hdpe = mean(hdpe / grand_total, na.rm = TRUE),
      avg_ldpe = mean(ldpe / grand_total, na.rm = TRUE),
      avg_pp = mean(pp / grand_total, na.rm = TRUE),
      avg_ps = mean(ps / grand_total, na.rm = TRUE),
      avg_pvc = mean(pvc / grand_total, na.rm = TRUE),
      avg_o = mean(o / grand_total, na.rm = TRUE),
      avg_forested = mean(forested_area, na.rm = TRUE),
      avg_threatened = mean(threatened_species, na.rm = TRUE)
    )
  leading_plastics <- country_stats |>
    pivot_longer(cols = !c(country, avg_forested, avg_threatened),
                 names_to = "plastic_type", values_to = "avg_value") |>
    slice_max(order_by = avg_value, n = 1, with_ties = FALSE) |>
    mutate(plastic_type = str_remove(plastic_type, "avg_"),
           plastic_type = str_to_upper(plastic_type),
           plastic_type = if_else(plastic_type == "O", "Other", plastic_type)) |>
    pull(plastic_type)
  country_stats |> mutate(leading_plastic = leading_plastics)
}

# precompute all needed data
top5_countries <- plastics |>
  group_by(country) |>
  summarise(ave_vol = first(volunteers) / n()) |>
  slice_max(ave_vol, n = 5) |> pull(country)

bot5_countries <- plastics |>
  group_by(country) |>
  summarise(ave_vol = first(volunteers) / n()) |>
  slice_min(ave_vol, n = 5) |> pull(country)

all_countries_rq2 <- population_plastics |>
  filter(parent_company == "Grand Total", grand_total > 0,
         !is.na(forested_area), !is.na(threatened_species)) |>
  distinct(country) |> pull(country)

plastic_ecology_summary_table <- map(all_countries_rq2, summarize_plastic_and_ecology) |>
  bind_rows()

top3_plastics <- population_plastics |>
  filter(parent_company == "Grand Total") |>
  summarize(across(c(pet, hdpe, ldpe, pp, ps, pvc), ~ sum(.x, na.rm = TRUE))) |>
  pivot_longer(everything(), names_to = "plastic", values_to = "total") |>
  slice_max(total, n = 3) |> pull(plastic)

gt_style <- function(gt_obj) {
  gt_obj |>
    tab_style(style = cell_fill(color = "#457B9D"), locations = cells_column_labels()) |>
    tab_style(style = cell_text(color = "white", weight = "bold"), locations = cells_column_labels()) |>
    tab_style(style = cell_fill(color = "#1D3557"), locations = cells_column_spanners()) |>
    tab_style(style = cell_text(color = "white", weight = "bold"), locations = cells_column_spanners())
}
```

# About

## Row

###

**Visualizing Plastics: Deliverable from Data**

This dashboard presents findings from a global plastic waste audit conducted by Break Free from Plastic (BFFP), covering 2019 and 2020 across 45+ countries. The data is observational and relies on self-reported volunteer submissions. We supplement the main dataset with country-level ecological and economic data from the API Ninjas Country API (accessed April 2025).

**Research Questions:**

1. Do countries with more volunteers tend to collect more proportions of both high and low density polyethylene plastic in 2019 and 2020?

2. Do countries with greater forest coverage and more threatened species show a different breakdown of plastic types?

**Data Sources:**

- **Main dataset:** Break Free from Plastic audit (TidyTuesday)
- **Supplementary dataset:** API Ninjas Country API

###

**Key Findings**

**RQ1 — Volunteers & Polyethylene:**

- No clear relationship between volunteer count and polyethylene proportion at the country level.
- A general positive trend exists — higher volunteers tend to correlate with more HDPE collected.
- Country-specific factors (laws, regulations) appear to be significant drivers.

**RQ2 — Ecology & Plastic Types:**

- PET drops from 71.7% → 38.3% → 21.5% as biodiversity level increases.
- Low forest coverage countries are overwhelmingly PET-dominated.
- Threatened species count is the only statistically significant predictor of PET proportion (p = 0.021, R² = 0.232).

# Research Question 1

## Row

### Tables {.tabset}

#### Table 1 — HDPE USA

```{r}
plastic_volunteer_summary("hdpe", "United States of America") |>
  gt() |>
  tab_header(title = md("**Table 1: HDPE Summary**"), subtitle = "United States of America") |>
  tab_spanner(label = "HDPE Metrics", columns = c(Ave_HDPE, Percent_HDPE, Total_Plastic)) |>
  tab_spanner(label = "Volunteer Metrics", columns = c(Ave_Vol, Total_Vol)) |>
  cols_label(Ave_HDPE = "Avg HDPE", Percent_HDPE = "% HDPE", Total_Plastic = "Total Plastic",
             Ave_Vol = "Avg Volunteers", Total_Vol = "Total Volunteers") |>
  gt_style()
```

Table 1 shows HDPE and volunteer statistics for the USA. The USA has 9.2% HDPE collected and a mean of 0.72 HDPE plastics across parent companies, with only 740 total volunteers.

#### Table 2 — LDPE USA

```{r}
plastic_volunteer_summary("ldpe", "United States of America") |>
  gt() |>
  tab_header(title = md("**Table 2: LDPE Summary**"), subtitle = "United States of America") |>
  tab_spanner(label = "LDPE Metrics", columns = c(Ave_LDPE, Percent_LDPE, Total_Plastic)) |>
  tab_spanner(label = "Volunteer Metrics", columns = c(Ave_Vol, Total_Vol)) |>
  cols_label(Ave_LDPE = "Avg LDPE", Percent_LDPE = "% LDPE", Total_Plastic = "Total Plastic",
             Ave_Vol = "Avg Volunteers", Total_Vol = "Total Volunteers") |>
  gt_style()
```

With the same volunteer count, the USA reports higher LDPE at 14.68% — more than 1 LDPE plastic collected per parent company on average.

#### Table 3 — HDPE Indonesia

```{r}
plastic_volunteer_summary("hdpe", "Indonesia") |>
  gt() |>
  tab_header(title = md("**Table 3: HDPE Summary**"), subtitle = "Indonesia") |>
  tab_spanner(label = "HDPE Metrics", columns = c(Ave_HDPE, Percent_HDPE, Total_Plastic)) |>
  tab_spanner(label = "Volunteer Metrics", columns = c(Ave_Vol, Total_Vol)) |>
  cols_label(Ave_HDPE = "Avg HDPE", Percent_HDPE = "% HDPE", Total_Plastic = "Total Plastic",
             Ave_Vol = "Avg Volunteers", Total_Vol = "Total Volunteers") |>
  gt_style()
```

Indonesia has 6,850 volunteers — nearly 10x the USA — yet only 4.25% HDPE. This suggests country-specific conditions beyond volunteer count drive plastic type composition.

#### Table 4 — Top & Bottom HDPE

```{r}
top5_hdpe <- map(top5_countries, ~plastic_volunteer_summary("hdpe", .x) |>
                   mutate(Country = .x, .before = 1)) |> list_rbind()
bot5_hdpe <- map(bot5_countries, ~plastic_volunteer_summary("hdpe", .x) |>
                   mutate(Country = .x, .before = 1)) |> list_rbind()

bind_rows(top5_hdpe |> mutate(Group = "Top 5", .before = 2),
          bot5_hdpe |> mutate(Group = "Bottom 5", .before = 2)) |>
  mutate(across(where(is.numeric), ~ifelse(is.nan(.), 0, .)),
         across(where(is.character), ~ifelse(. == "NaN%", "0%", .))) |>
  gt() |>
  tab_header(title = md("**Table 4: HDPE — Top & Bottom 5 Volunteer Countries**")) |>
  tab_spanner(label = "Location", columns = c(Country, Group)) |>
  tab_spanner(label = "HDPE Metrics", columns = c(Ave_HDPE, Percent_HDPE, Total_Plastic)) |>
  tab_spanner(label = "Volunteer Metrics", columns = c(Ave_Vol, Total_Vol)) |>
  cols_label(Ave_HDPE = "Avg HDPE", Percent_HDPE = "% HDPE", Total_Plastic = "Total Plastic",
             Ave_Vol = "Avg Volunteers", Total_Vol = "Total Volunteers") |>
  gt_style()
```

No clear pattern emerges. Taiwan, Bhutan, and Benin report 0% HDPE despite being top 5 volunteer countries, while Cote D'Ivoire reports 100% HDPE.

#### Table 5 — Top & Bottom LDPE

```{r}
top5_ldpe <- map(top5_countries, ~plastic_volunteer_summary("ldpe", .x) |>
                   mutate(Country = .x, .before = 1)) |> list_rbind()
bot5_ldpe <- map(bot5_countries, ~plastic_volunteer_summary("ldpe", .x) |>
                   mutate(Country = .x, .before = 1)) |> list_rbind()

bind_rows(top5_ldpe |> mutate(Group = "Top 5", .before = 2),
          bot5_ldpe |> mutate(Group = "Bottom 5", .before = 2)) |>
  mutate(across(where(is.numeric), ~ifelse(is.nan(.), 0, .)),
         across(where(is.character), ~ifelse(. == "NaN%", "0%", .))) |>
  gt() |>
  tab_header(title = md("**Table 5: LDPE — Top & Bottom 5 Volunteer Countries**")) |>
  tab_spanner(label = "Location", columns = c(Country, Group)) |>
  tab_spanner(label = "LDPE Metrics", columns = c(Ave_LDPE, Percent_LDPE, Total_Plastic)) |>
  tab_spanner(label = "Volunteer Metrics", columns = c(Ave_Vol, Total_Vol)) |>
  cols_label(Ave_LDPE = "Avg LDPE", Percent_LDPE = "% LDPE", Total_Plastic = "Total Plastic",
             Ave_Vol = "Avg Volunteers", Total_Vol = "Total Volunteers") |>
  gt_style()
```

4 out of 5 top volunteer countries report 0% LDPE. No clear positive relationship between volunteers and LDPE proportion.

### Visualizations {.tabset}

#### Figure 1 — Bubble Plot

```{r}
#| fig-alt: "Bubble plot showing average volunteers vs HDPE percentage by country on log scale"
all_countries_rq1 <- plastics |> distinct(country) |> pull(country)
bubble_data <- map(all_countries_rq1, ~plastic_volunteer_summary("hdpe", .x) |>
                     mutate(Country = .x)) |>
  list_rbind() |>
  mutate(Percent_HDPE = as.numeric(str_remove(Percent_HDPE, "%"))) |>
  filter(Percent_HDPE < 100)

p <- ggplot(bubble_data, aes(x = Ave_Vol, y = Percent_HDPE,
                        size = Total_Plastic, color = Country,
                        text = paste("Country:", Country,
                                     "<br>Avg Vol:", Ave_Vol,
                                     "<br>% HDPE:", Percent_HDPE,
                                     "<br>Total:", comma(Total_Plastic)))) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(range = c(2, 15)) +
  scale_x_log10() +
  scale_color_viridis_d() +
  guides(color = "none") +
  theme_minimal() +
  labs(title = "Figure 1: Average Volunteers vs. HDPE Plastic (%) by Country",
       subtitle = "Bubble size represents total plastic collected",
       x = "Average Volunteers (log scale)", y = "% HDPE", size = "Total Plastic")

ggplotly(p, tooltip = "text")
```

#### Figure 2 — Dumbbell Plot

```{r}
#| warning: false
#| fig-height: 6
#| fig-alt: "Dumbbell plot comparing HDPE and LDPE for top and bottom 10 volunteer countries"
plastics |>
  filter(year %in% c(2019, 2020), !country %in% c("EMPTY", "Grand Total")) |>
  mutate(country = case_when(
    str_to_upper(str_trim(country)) == "NIGERIA" ~ "Nigeria",
    str_to_upper(str_trim(country)) == "TAIWAN_ REPUBLIC OF CHINA (ROC)" ~ "Taiwan",
    TRUE ~ str_to_title(str_trim(country))),
    hdpe_prop = hdpe / grand_total, ldpe_prop = ldpe / grand_total) |>
  group_by(country) |>
  summarise(hdpe_prop = mean(hdpe_prop, na.rm = TRUE),
            ldpe_prop = mean(ldpe_prop, na.rm = TRUE),
            total_volunteers = sum(volunteers, na.rm = TRUE)) |>
  arrange(total_volunteers) |>
  slice(c(1:10, (n()-9):n())) |>
  mutate(country = fct_reorder(country, total_volunteers),
         volunteer_group = c(rep("Bottom 10", 10), rep("Top 10", 10))) |>
  ggplot(aes(y = country)) +
  geom_segment(aes(x = hdpe_prop, xend = ldpe_prop, yend = country), color = "gray70") +
  geom_point(aes(x = hdpe_prop), color = "steelblue", size = 2) +
  geom_point(aes(x = ldpe_prop), color = "coral", size = 2) +
  facet_wrap(~ volunteer_group, scales = "free_y") +
  labs(title = "Figure 2: HDPE vs LDPE Proportion by Country",
       subtitle = "Top 10 and Bottom 10 Countries by Total Volunteer Count",
       x = "Proportion of Total Plastic", y = NULL,
       caption = "Blue = HDPE | Orange = LDPE") +
  theme_minimal()
```

# Research Question 2

## Row

### Tables {.tabset}

#### Table 6 — Forest Coverage

```{r}
plastic_ecology_summary_table |>
  mutate(forest_coverage_level = cut(avg_forested, breaks = c(0, 19, 36, 100),
                            labels = c("Low (< 19%)", "Medium (19-36%)", "High (> 36%)"))) |>
  group_by(forest_coverage_level) |>
  slice_min(avg_forested, n = 5) |>
  ungroup() |>
  arrange(avg_forested) |>
  gt(groupname_col = "forest_coverage_level") |>
  tab_header(title = "Table 6: Plastic Type Proportions by Country and Forest Coverage",
             subtitle = "5 least forested countries within each coverage level") |>
  cols_label(country = "Country", avg_pet = "PET", avg_hdpe = "HDPE", avg_ldpe = "LDPE",
             avg_pp = "PP", avg_ps = "PS", avg_pvc = "PVC", avg_o = "Other",
             avg_forested = "Forested (%)", avg_threatened = "Threatened Species",
             leading_plastic = "Leading Plastic") |>
  cols_move(columns = leading_plastic, after = country) |>
  tab_spanner(label = "Average Plastic Proportion", columns = avg_pet:avg_o) |>
  fmt_percent(columns = avg_pet:avg_o, decimals = 1) |>
  fmt_number(columns = avg_forested, decimals = 1) |>
  fmt_number(columns = avg_threatened, decimals = 0) |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_spanners()) |>
  tab_style(style = cell_fill(color = "#18BC9C"), locations = cells_row_groups(groups = "High (> 36%)")) |>
  tab_style(style = cell_fill(color = "#F39C12"), locations = cells_row_groups(groups = "Medium (19-36%)")) |>
  tab_style(style = cell_fill(color = "#E74C3C"), locations = cells_row_groups(groups = "Low (< 19%)")) |>
  tab_style(style = cell_text(weight = "bold", color = "#2C3E50"),
            locations = cells_body(columns = leading_plastic)) |>
  opt_stylize(style = 6, color = "gray") |>
  tab_options(column_labels.background.color = "#2C3E50",
              heading.background.color = "#2C3E50") |>
  sub_missing(missing_text = "—") |>
  cols_hide(forest_coverage_level)
```

Low coverage countries are overwhelmingly PET-dominated (Maldives, Tunisia: 100% PET). As coverage increases, plastic composition diversifies — Cameroon leads with LDPE (39.8%) in the high coverage group.

#### Table 7 — Biodiversity Level

```{r}
population_plastics |>
  filter(parent_company == "Grand Total", grand_total > 0,
         !is.na(threatened_species), !is.na(forested_area)) |>
  mutate(pet_prop = pet/grand_total, hdpe_prop = hdpe/grand_total,
         ldpe_prop = ldpe/grand_total, pp_prop = pp/grand_total,
         ps_prop = ps/grand_total, pvc_prop = pvc/grand_total,
         o_prop = o/grand_total,
         threatened_species_level = cut(threatened_species,
                       breaks = c(0, 131, 556, Inf),
                       labels = c("Low (< 131)", "Medium (131-556)", "High (> 556)"))) |>
  group_by(threatened_species_level) |>
  summarize(n_countries = n_distinct(country),
            avg_pet = mean(pet_prop, na.rm = TRUE),
            avg_hdpe = mean(hdpe_prop, na.rm = TRUE),
            avg_ldpe = mean(ldpe_prop, na.rm = TRUE),
            avg_pp = mean(pp_prop, na.rm = TRUE),
            avg_ps = mean(ps_prop, na.rm = TRUE),
            avg_pvc = mean(pvc_prop, na.rm = TRUE),
            avg_o = mean(o_prop, na.rm = TRUE),
            avg_forested = mean(forested_area, na.rm = TRUE)) |>
  gt() |>
  tab_header(title = "Table 7: Plastic Type Composition by Biodiversity Level") |>
  cols_label(threatened_species_level = "Biodiversity Level", n_countries = "Countries",
             avg_pet = "PET", avg_hdpe = "HDPE", avg_ldpe = "LDPE", avg_pp = "PP",
             avg_ps = "PS", avg_pvc = "PVC", avg_o = "Other",
             avg_forested = "Avg. Forested (%)") |>
  fmt_percent(columns = avg_pet:avg_o, decimals = 1) |>
  fmt_number(columns = avg_forested, decimals = 1) |>
  tab_spanner(label = "Average Plastic Proportion", columns = avg_pet:avg_o) |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_spanners()) |>
  tab_style(style = cell_fill(color = "#18BC9C"),
            locations = cells_body(columns = threatened_species_level,
                                   rows = threatened_species_level == "High (> 556)")) |>
  tab_style(style = cell_fill(color = "#F39C12"),
            locations = cells_body(columns = threatened_species_level,
                                   rows = threatened_species_level == "Medium (131-556)")) |>
  tab_style(style = cell_fill(color = "#E74C3C"),
            locations = cells_body(columns = threatened_species_level,
                                   rows = threatened_species_level == "Low (< 131)")) |>
  tab_options(column_labels.background.color = "#2C3E50",
              heading.background.color = "#2C3E50") |>
  sub_missing(missing_text = "—")
```

PET drops from 71.7% → 38.3% → 21.5% as biodiversity increases — a drop of over 50 percentage points. HDPE rises from 1.5% to 12.4% across low to high biodiversity countries.

### Visualizations {.tabset}

#### Figure 3 — Scatter Plot

```{r}
#| message: false
#| warning: false
#| fig-height: 6
#| fig-alt: "Scatter plot of forested area vs threatened species count colored by leading plastic type"
plot_data <- plastic_ecology_summary_table |>
  mutate(leading_plastic = fct_infreq(leading_plastic) |> fct_relevel("Other", after = Inf),
         point_alpha = if_else(leading_plastic == "Other", 0.4, 0.9))

ggplot(plot_data, aes(x = avg_forested, y = avg_threatened,
                      color = leading_plastic, label = country)) +
  geom_point(aes(alpha = point_alpha), size = 1.5) +
  geom_text_repel(size = 2, max.overlaps = Inf, box.padding = 0.25,
                  point.padding = 0.15, segment.color = "gray70",
                  segment.size = 0.25, show.legend = FALSE) +
  scale_y_log10(labels = scales::label_comma()) +
  scale_alpha_identity() +
  scale_color_manual(values = c("PET" = "#E74C3C", "HDPE" = "#3498DB",
                                "LDPE" = "#F39C12", "PP" = "#18BC9C",
                                "PS" = "#9B59B6", "PVC" = "#2C3E50",
                                "Other" = "#95A5A6")) +
  labs(x = "Forested Area (%)", y = "Threatened Species (log scale)",
       color = "Leading Plastic",
       title = "Figure 3: Where Do Different Plastic Types Dominate?") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))
```

PET countries (red) cluster in the lower left — low forest, few threatened species. PP and LDPE dominate in higher ecological richness zones. UK is a notable outlier with PS despite low forest coverage.

#### Figure 4 — Ridge Plot

```{r}
#| message: false
#| fig-height: 5
#| fig-alt: "Ridge plot showing threatened species distribution across top 3 plastic types"
population_plastics |>
  filter(parent_company == "Grand Total", grand_total > 0, !is.na(threatened_species)) |>
  pivot_longer(cols = all_of(top3_plastics), names_to = "plastic_type", values_to = "count") |>
  filter(count > 0) |>
  mutate(plastic_type = factor(toupper(plastic_type), levels = toupper(top3_plastics))) |>
  ggplot(aes(x = threatened_species, y = plastic_type, fill = plastic_type, weight = count)) +
  geom_density_ridges(scale = 0.8, alpha = 0.7) +
  scale_x_log10(labels = scales::label_comma()) +
  scale_fill_manual(values = c("#3498DB", "#F39C12", "#18BC9C")) +
  labs(title = "Figure 4: Threatened Species Counts Across Top 3 Plastic Types",
       subtitle = "Each country weighted by how much of that plastic it collected",
       x = "Threatened Species Count (log scale)", y = "Plastic Type") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11),
        legend.position = "none")
```

The PP ridge shifts furthest right — PP is collected in more biodiverse countries. PET centers around 100-500 threatened species, confirming it dominates in less ecologically rich nations.

# Data Model

## Row

### Model Results

```{r}
pet_model <- population_plastics |>
  filter(parent_company == "Grand Total", grand_total > 0,
         !is.na(forested_area), !is.na(threatened_species),
         !is.na(gdp_per_capita), !is.na(volunteers)) |>
  mutate(pet_prop = pet / grand_total) |>
  lm(pet_prop ~ forested_area + log(threatened_species) +
                log(gdp_per_capita) + volunteers, data = _)

anova(pet_model) |>
  tidy() |>
  mutate(pct_variance = sumsq / sum(sumsq) * 100,
         term = recode(term, "volunteers" = "Volunteers",
                       "forested_area" = "Forested Area",
                       "log(threatened_species)" = "Threatened Species (log)",
                       "log(gdp_per_capita)" = "GDP Per Capita (log)",
                       "Residuals" = "Unexplained")) |>
  select(term, df, sumsq, pct_variance, statistic, p.value) |>
  gt() |>
  tab_header(title = "Table 8: What Predicts PET Plastic Proportion?",
             subtitle = "Variance partitioned across ecological, economic and volunteer predictors") |>
  cols_label(term = "Source", df = "df", sumsq = "Sum of Squares",
             pct_variance = "% of Variance", statistic = "F", p.value = "p-value") |>
  fmt_number(columns = c(sumsq, statistic), decimals = 3) |>
  fmt_number(columns = pct_variance, decimals = 1) |>
  fmt_number(columns = p.value, decimals = 4) |>
  sub_missing(missing_text = "—") |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) |>
  tab_style(style = list(cell_fill(color = "#18BC9C"), cell_text(weight = "bold")),
            locations = cells_body(rows = p.value < 0.05)) |>
  tab_options(column_labels.background.color = "#2C3E50",
              heading.background.color = "#2C3E50") |>
  tab_source_note("Note: PET modeled as it showed the strongest relationship with ecological variables.") |>
  tab_source_note(paste0("Model R² = ", round(summary(pet_model)$r.squared, 3),
                         " · n = ", nobs(pet_model)))
```

### Interpretation

**Key Finding**

Threatened species count is the only statistically significant predictor of PET plastic proportion (F = 5.841, p = 0.0209), explaining 12.5% of variance. As threatened species increases, PET proportion decreases (coefficient = -0.1325).

Forested area showed a similar negative trend (p = 0.0794) but was not statistically significant. Together the model explains about 23% of variation in PET proportion.

Circling back to our research question — forested area percentage and number of threatened species do affect the distribution of plastic, particularly PET proportion. Countries with greater ecological richness collect fundamentally different plastic waste compositions.